Aldebaran~b's temperate past uncovered in planet search data

### 2018-02-05 paper/aldebaran.tex
Final edits to manuscript before submission.

Radial velocity variations intrinsic to the star, such as those caused by stellar oscillations, have previously limited the precision with which red giant planets have been studied \citep{2005PASJ...57...97S}. We aim to model out these noise processes and use them to recover asteroseismic information, as well as to improve our estimates of the orbital parameters of the planet. It is easy to observe Aldebaran and similarly-bright stars with ground-based spectroscopic instruments, typically only requiring short exposures that can be obtained even under adverse observing conditions. There is indeed a considerable archive of such observations already, as a legacy of radial velocity (RV) surveys conducted to find exoplanets. In most cases, however, these have not so far been useful for asteroseismology because these RV data are sparsely and irregularly sampled. Because we have to pause observations during the day, during poor weather conditions, or simply when targets of higher priority are being observed, we get time series which may have significant and uneven gaps. This introduces a window-function effect: the power spectrum as constructed for example by a Fourier transform, or a Lomb-Scargle Periodogram \citep{lomb,scargle}, is convolved with the Fourier transform of the window function, introducing strong sidelobes adjacent to real frequency peaks and causing crosstalk between adjacent frequency channels. This imposes significant limitations both on the signal-to-noise and frequency resolution of power spectra derived from linear methods such as the Lomb-Scargle periodogram, and in practice makes asteroseismology with these conventional approaches difficult or impossible from the ground for stars with oscillation periods ranging from $\sim 12$~h to $\sim$~a few days.

@@

\citet{Hatzes2015} reported on RV observations of Aldebaran from the coud\'{e}
\'{e}chelle spectrograph of the Th\"{u}ringer Landessternwarte Tautenburg (TLS),
the Tull Spectrograph of the McDonald 2.7 m telescope, and the Bohyunsan
Observatory \'{E}chelle Spectrograph (BOES) spectrograph of the Bohyunsan
Optical Astronomy Observatory (BOAO) spanning the period from 2000.01 to
2013.92.  Combined with earlier observations \citep{hatzes1993}, we construct a
full time series spanning more than two decades from 1980.80 to 2013.71
(Figure~\ref{alldata}) and fit this with a Keplerian model and a CARMA process.
We find evidence for a low-quality ($Q = \HatzesQRange{}$) oscillatory mode with
$\nu_\mathrm{max} = \HatzesNuMaxRange{} \, \mu\mathrm{Hz}$.
This is consistent with the presence of a number of un-resolved asteroseismic modes in the RV data.
The low Q factor in this case refers to the inverse fractional width of the band over which the modes are excited, not the quality of any individual mode.
Our posterior for \numax{} from this data set is shown in Figure
\ref{fig:numax-datasets}.

@@

\subsection{SONG Observations}    % this goes into section 4.1
\label{sec:SONG-observations} In order to confirm the oscillations detected in
the archival data, we used the Hertzsprung SONG telescope
\citep{2017ApJ...836..142G} to conduct high time cadence follow-up observations.
These were carried out in the highest resolution mode ($R\approx110000$) with an
integration time of 30\,s, between 2016 September 27 and 2017 January 12. During
this time we attempted to obtain at least one visit per available night, for a
total of 254 epochs over the campaign. The spectra were extracted using the SONG
pipeline \citep[see][]{2017ApJ...836..142G}. SONG employs an iodine cell for
precise wavelength calibration and the velocities were determined using the
\textit{iSONG} software following the same procedures as described in
\cite{2017ApJ...836..142G}. The typical velocity precision per visit is
2.5$\,\rm m\,s^{-1}$ allowing us to easily detect the oscillations which
display a peak-to-peak amplitude of $\approx \, 170 \, \rm m \, s^{-1}$.
Figure~\ref{songdata} displays the radial velocities obtained -- oscillations
are easily visible as well as a long-term trend, probably due to the planetary
companion.  From this data set alone we also find evidence for a low-quality ($Q
= \SONGQRange{}$) oscillatory mode with $\nu_\mathrm{max} = \SONGNuMaxRange{} \,
\mu\mathrm{Hz}$ which is consistent with the inference from the archival data.
This posterior is shown in Figure \ref{fig:numax-datasets}.

@@

We have determined the stellar properties
of Aldebaran with the additional constraint of \numax determined from this
analysis.  The details of the stellar modelling are presented in Appendix
\ref{sec:mod}.  We have considered the impact of our additional \numax
constraint and run our analysis for different sets of spectroscopic estimates.
For final values we adopt the \citet{2012Sheffield} spectroscopic solution as
being `middle of the road' estimates.  For the analysis without constraint on
\numax we find estimates of the stellar properties as $M = 1.27^{+0.24}_{-0.20}
\, \mathrm{M_{\odot}}$ and age $4.9^{+3.6}_{-2.0} \, \rm Gyr$.  With the
inclusion of \numax we find $M = 1.16\, \pm \, {0.07} \, \mathrm{M_{\odot}}$ and
age $6.4^{+1.4}_{-1.1} \, \rm Gyr$.  It is clear that the addition of \numax
provides substantially more precise estimates of mass and age. With this stellar
mass, the radial velocity signal translates to an inclination-dependent
planetary mass of $M\,\sin\,{i} = 5.8 \pm 0.7~M_J$
\citep[Equation~1]{Torres2008}, a massive giant planet; for sufficiently high
inclinations ($\gtrsim 45^\circ$), the mass of Aldebaran~b may exceed $13~M_J$,
making it a possible low-mass brown dwarf candidate. A recent study by \citet{2018arXiv180105239H} has cast doubt on the validity of the detection of the massive planet $\gamma$~Draconis~b, and by extension similar planets around similar K~giants such as Aldebaran. We remain convinced that the $\gamma$~Draconis problem is not an issue in this case, and Aldebaran~b is a bona fide planet, for reasons outlined in Appendix~\ref{gamma_dra}.

@@

With CARMA models we have confirmed the previously-suspected planet around Aldebaran using archival data, and additionally detected acoustic oscillations that permit a mass determination with a precision of $\sim 6\%$. We have confirmed both of these results with new ground-based observations, and space-based photometry with \ktwo.

@@

The Kalman filter implementation of the CARMA likelihood is not unique.
\citet{Foreman-Mackey2017} show how a dimensional expansion
can be used to reduce the standard GP covariance matrix for this
same process,  with autocorrelation given by Eq.\ \eqref{eq:CARMA-ACF} under
observations at times $t_k$, to a banded form that can then be
Cholesky-decomposed in the standard GP likelihood function in
$\mathcal{O}\left( n \right)$ time. These `\emph{celerite}' models are formally equivalent to the Kalman CARMA model and differ only in their computational implementation.

@@

The \kepler Space Telescope \citep{2010sci...327..977b}
suffered a critical reaction wheel failure in May 2013, which made it impossible to maintain a stable pointing and therefore continue its nominal mission. It was revived as \ktwo \citep{howell14}, balanced by orienting perpendicular to the Sun. This requires that \ktwo observes fields in the Ecliptic in $\sim80$~d Campaigns, of which Aldebaran was observed in Campaign~13.

>>>>

Radial velocity variations intrinsic to the star, such as those caused by stellar oscillations, have previously limited the precision with which red giant planets have been studied \citep{2005PASJ...57...97S}. We aim to model these noise processes and use them to recover asteroseismic information, as well as to improve our estimates of the orbital parameters of the planet. It is easy to observe Aldebaran and similarly-bright stars with ground-based spectroscopic instruments, typically only requiring short exposures that can be obtained even under adverse observing conditions. There is indeed a considerable archive of such observations already, as a legacy of radial velocity (RV) surveys conducted to find exoplanets. In most cases, however, these have not so far been useful for asteroseismology because these RV data are sparsely and irregularly sampled. Because we have to pause observations during the day, during poor weather conditions, or simply when targets of higher priority are being observed, we get time series which may have significant and uneven gaps. This introduces a window-function effect: the power spectrum as constructed for example by a Fourier transform, or a Lomb-Scargle Periodogram \citep{lomb,scargle}, is convolved with the Fourier transform of the window function, introducing strong sidelobes adjacent to real frequency peaks and causing crosstalk between adjacent frequency channels. This imposes significant limitations both on the signal-to-noise and frequency resolution of power spectra derived from linear methods such as the Lomb-Scargle periodogram, and in practice makes asteroseismology with these conventional approaches difficult or impossible from the ground for stars with oscillation periods ranging from $\sim 12$~h to $\sim$~a few days.

@@

\citet{Hatzes2015} reported on RV observations of Aldebaran from the coud\'{e}
\'{e}chelle spectrograph of the Th\"{u}ringer Landessternwarte Tautenburg (TLS),
the Tull Spectrograph of the McDonald 2.7 m telescope, and the Bohyunsan
Observatory \'{E}chelle Spectrograph (BOES) spectrograph of the Bohyunsan
Optical Astronomy Observatory (BOAO) spanning the period from 2000.01 to
2013.92.  Combined with earlier observations \citep{hatzes1993}, we construct a
full time series spanning more than two decades from 1980.80 to 2013.71
(Figure~\ref{alldata}) and fit this with a Keplerian model and a CARMA process.
We find evidence for a low-quality ($Q = \HatzesQRange{}$) oscillatory mode with
$\nu_\mathrm{max} = \HatzesNuMaxRange{} \, \mu\mathrm{Hz}$ (here and throughout
we quote the posterior median and the range from the 0.16 to the 0.84 posterior
quantile).  This is consistent with the presence of a number of un-resolved
asteroseismic modes in the RV data.  The low Q factor in this case refers to the
inverse fractional width of the band over which the modes are excited, not the
quality of any individual mode.  Our posterior for \numax{} from this data set
is shown in Figure \ref{fig:numax-datasets}.

@@

\subsection{SONG Observations}    % this goes into section 4.1
\label{sec:SONG-observations} In order to confirm the oscillations detected in
the archival data, we used the Hertzsprung SONG telescope
\citep{2017ApJ...836..142G} to conduct high time cadence follow-up observations.
These were carried out in the highest resolution mode ($R\approx110000$) with an
integration time of 30\,s, between 2016 September 27 and 2017 January 12. During
this time we attempted to obtain at least one visit per available night, for a
total of 254 epochs over the campaign. The spectra were extracted using the SONG
pipeline \citep[see][]{2017ApJ...836..142G}. SONG employs an iodine cell for
precise wavelength calibration and the velocities were determined using the
\textit{iSONG} software following the same procedures as described in
\cite{2017ApJ...836..142G}. The typical velocity precision per visit is
2.5$\,\rm m\,s^{-1}$ allowing us to easily detect the oscillations which display
a peak-to-peak amplitude of $\approx \, 170 \, \rm m \, s^{-1}$.
Figure~\ref{songdata} displays the radial velocities obtained---oscillations
are easily visible as well as a long-term trend, due to the planetary companion.
From this data set alone we also find evidence for a low-quality ($Q =
\SONGQRange{}$) oscillatory mode with $\nu_\mathrm{max} = \SONGNuMaxRange{} \,
\mu\mathrm{Hz}$ which is consistent with the inference from the archival data.
This posterior is shown in Figure \ref{fig:numax-datasets}.

@@

We have used the additional constraint of \numax determined from this analysis
to update the stellar properties of Aldebaran.  The details of the stellar
modelling are presented in Appendix \ref{sec:mod}.  We have considered the
impact of our additional \numax constraint and run our analysis for different
sets of spectroscopic estimates.  For final values we adopt the
\citet{2012Sheffield} spectroscopic solution as being `middle of the road'
estimates.  For the analysis without constraint on \numax we find estimates of
the stellar properties as $M = 1.27^{+0.24}_{-0.20} \, \mathrm{M_{\odot}}$ and
age $4.9^{+3.6}_{-2.0} \, \rm Gyr$.  With the inclusion of \numax we find $M =
1.16\, \pm \, {0.07} \, \mathrm{M_{\odot}}$ and age $6.4^{+1.4}_{-1.1} \, \rm
Gyr$.  It is clear that the addition of \numax provides substantially more
precise estimates of mass and age. With this stellar mass, the radial velocity
signal translates to an inclination-dependent planetary mass of $M\,\sin\,{i} =
5.8 \pm 0.7~M_J$ \citep[Equation~1]{Torres2008}, a massive giant planet; for
sufficiently high inclinations ($\gtrsim 45^\circ$), the mass of Aldebaran~b may
exceed $13~M_J$, making it a possible low-mass brown dwarf candidate. A recent
study by \citet{2018arXiv180105239H} has cast doubt on the validity of the
detection of the massive planet $\gamma$~Draconis~b, and by extension similar
planets around similar K~giants such as Aldebaran. We remain convinced that the
$\gamma$~Draconis problem is not an issue in this case, and Aldebaran~b is a
bona fide planet, for reasons outlined in Appendix~\ref{gamma_dra}.

@@

Using sophisticated time-domain CARMA models for the stellar activity in
Aldebaran, we have confirmed the previously-suspected planet and additionally
detected acoustic oscillations that permit a mass determination with a precision
of $\sim 6\%$. We have confirmed both of these results with new ground-based
observations, and space-based photometry with \ktwo.

@@

The Kalman filter implementation of the CARMA likelihood is not unique.
\citet{Foreman-Mackey2017} show how a dimensional expansion can be used to
reduce the standard GP covariance matrix for this same process,  with
autocorrelation given by Eq.\ \eqref{eq:CARMA-ACF} under observations at times
$t_k$, to a banded form that can then be Cholesky-decomposed in the standard GP
likelihood function in $\mathcal{O}\left( n \right)$ time. These
`\emph{celerite}' models are formally equivalent to a Kalman CARMA model with $q
= p-1$ and differ only in their computational implementation.

@@

The \kepler Space Telescope \citep{2010sci...327..977b}
suffered a critical reaction wheel failure in May 2013, which made it impossible to maintain a stable pointing and therefore continue its nominal mission. It was revived as \ktwo \citep{howell14}, balanced by orienting perpendicular to the Sun. This requires that \ktwo observes fields in the Ecliptic in $\sim80$~d Campaigns; Aldebaran was observed in Campaign~13.
